A Software Engineering Perspective on \\ OpenMP and PThreads

### 2017-10-16 sections/04-research-design.tex
Edits and changes to make the camera ready version.

We devised an assignment to measure the students' difficulties to learn and use
\textit{Pthreads} and \textit{OpenMP} to solve a simple problem. Students had
to implement parallel versions of a sequential code using \textit{Pthreads} and
\textit{OpenMP}.  We provided a sequential implementation for the calculation
of the Mandelbrot Set~\cite{douady1984etude}. The sequential code presented an
embarrassingly parallel problem with two nested loops, and a third loop with
dependent iterations that could not be parallelized easily.

@@

The questionnaire had two parts. The first part asked questions
about the students' previous experiences with common programming languages and
parallel and distributed programming concepts and tools. We were also
interested in students' perception of the impact of our classes on
\textit{Pthreads} and \textit{OpenMP} on their learning difficulty.

>>>>

We devised an assignment to measure the students' difficulties to learn and use
\textit{Pthreads} and \textit{OpenMP} to solve a simple problem. Students had
to form groups of 2 to 3 members and implement parallel versions of a
sequential code using \textit{Pthreads} and \textit{OpenMP}.  We provided a
sequential implementation for the calculation of the Mandelbrot
Set~\cite{douady1984etude}. The sequential code presented an embarrassingly
parallel problem with two nested loops, and a third loop with dependent
iterations that could not be parallelized easily.

@@

After students organized themselves in groups of 2 to 3 students and designed
and implemented their parallel versions of the code, each individual student
was asked to answer a questionnaire. The questionnaire had two parts. The first
part asked questions about the students' previous experiences with common
programming languages and parallel and distributed programming concepts and
tools. We were also interested in students' perception of the impact of our
classes on \textit{Pthreads} and \textit{OpenMP} on their learning difficulty.

### 2017-10-16 sections/05-responses.tex
Edits and changes to make the camera ready version.

We asked the students to answer 3 sets of questions regarding their
experience with \textit{OpenMP} and \textit{Pthreads}, \textit{after}
completing the assignment.
The first set of questions used a scale defined by us with 5 levels to measure the relationship
of the students with the two APIs in the context of the assignment. This kind of scale enables respondents to express their agreement
or disagreement with specific statements~\cite{likert1932technique}. Respondents choose between five
options: \textit{Strongly Agree} (SA), \textit{Agree} (A), \textit{Neutral}
(N), \textit{Disagree} (D) and \textit{Strongly Disagree}.

@@

\begin{table}[htpb]
    \centering
    \begin{tabular}{@{}p{0.7\columnwidth}p{0.1\columnwidth}p{0.05\columnwidth}@{}}
        \toprule
        \multicolumn{2}{c}{\scriptsize{It is easy to$\dots$}} & \textnumero \\ \midrule
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Parallelize loops with independent iterations using:}}} & \scriptsize{OpenMP} & $(1)$ \\
        & \scriptsize{Pthreads} & $(2)$ \\
        \addlinespace{}
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Parallelize nested loops with independent iterations using:}}} & \scriptsize{OpenMP} & $(3)$ \\
        &  \scriptsize{Pthreads} & $(4)$ \\
        \addlinespace{}
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Improve the performance of sequential code using:}}} & \scriptsize{OpenMP} & $(5)$  \\
        &  \scriptsize{Pthreads} & $(6)$ \\ \bottomrule
    \end{tabular}
    \caption{Scale questions for Figure \ref{fig:likert}}
    \label{tab:likert}
    \vspace{-0.6em}
\end{table}

\begin{figure}[htpb]
    \vspace{-2em}
    \centering
    \includegraphics[width=0.85\columnwidth]{likert_questions}
    \caption{Students' responses to the scale questions in Table \ref{tab:likert}}
    \label{fig:likert}
    \vspace{-1em}
\end{figure}

>>>>

We asked the students to answer 3 sets of questions regarding their experience
with \textit{OpenMP} and \textit{Pthreads}, \textit{after} completing the
assignment.  The first set of questions used a Likert
scale~\cite{likert1932technique} to measure the relationship of the students
with the two APIs in the context of the assignment. The Likert scale enables
respondents to express their agreement or disagreement with specific
statements. Respondents choose between five options: \textit{Strongly Agree}
(SA), \textit{Agree} (A), \textit{Neutral} (N), \textit{Disagree} (D) and
\textit{Strongly Disagree}.

@@

\begin{table}[htpb]
    \centering
    \begin{tabular}{@{}p{0.7\columnwidth}p{0.1\columnwidth}p{0.05\columnwidth}@{}}
        \toprule
        \multicolumn{2}{c}{\scriptsize{It is easy to$\dots$}} & \textnumero \\ \midrule
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Parallelize loops with independent iterations using:}}} & \scriptsize{OpenMP} & $(1)$ \\
        & \scriptsize{Pthreads} & $(2)$ \\
        \addlinespace{}
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Parallelize nested loops with independent iterations using:}}} & \scriptsize{OpenMP} & $(3)$ \\
        &  \scriptsize{Pthreads} & $(4)$ \\
        \addlinespace{}
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Improve the performance of sequential code using:}}} & \scriptsize{OpenMP} & $(5)$  \\
        &  \scriptsize{Pthreads} & $(6)$ \\ \bottomrule
    \end{tabular}
    \caption{Likert scale questions for Figure \ref{fig:likert}}
    \label{tab:likert}
    \vspace{-0.6em}
\end{table}

\begin{figure}[htpb]
    \vspace{-2em}
    \centering
    \includegraphics[width=0.85\columnwidth]{likert_questions}
    \caption{Students' responses to the Likert scale questions in Table \ref{tab:likert}}
    \label{fig:likert}
    \vspace{-1em}
\end{figure}

### 2017-10-16 sections/07-conclusion.tex
Edits and changes to make the camera ready version.

>>>>

Threats to the validity of this study include it being performed in a single
class of the Concurrent, Parallel and Distributed Computing course.  A
statistically relevant change to this study would be collecting data on
multiple instances of the course over a period of multiple semesters. The
subjectivity of the questionnaire and of the experience of the students is also
a threat to validity. Additionally, we did not control for the order in which
students performed their implementations in \textit{Pthreads} and
\textit{OpenMP}, and we did not had separate groups learning each technology.
